---
title: "Skills for Analysis Assignment 6"
author: "GSD SES 5394"
date: "Spring 2023"
output: 
  rmdformats::material
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This page presents an approach to estimating a trip production model based on data from the National Household Travel Survey. 

The examples here use the following packages:

```{r, message=FALSE, results='hide'}
library(here)
library(tidyverse)
library(downloader)
```

# Download data from the National Household Travel Survey

You can download the full 2017 NHTS datasets using the code
below, or you can just navigate your browser to the url, 
download the zipped file from there, and then extract the files
to the directory where you want to save them
(that would probably be faster).

```{r, eval=FALSE}
url <- "https://nhts.ornl.gov/assets/2016/download/csv.zip"
nhts_zipped <- here("nhts.zip")
nhts_dir <- here("nhts")

download(url, nhts_zipped, mode="wb") 
unzip(nhts_zipped, exdir = nhts_dir)
```

The extracted directory will have the following files:

* Citation.docx
* hhpub.csv
* perpub.csv
* trippub.csv
* vehpub.csv

# Load trips table and filter to study area

You will likely be primarily working with the trips file
(trippub.csv). You will want to load this to R and filter to 
only include trips by households in your study area.

Use the interactive codebook browser to find the code for your
study area (the variable will be HH_CBSA).

```{r, eval=FALSE}
okc_trips <- here("nhts",
                  "trippub.csv") %>%
  read_csv() %>%
  filter(HH_CBSA == "36420")
```

Core-based statistical areas with populations less than one 
million will not be specified in the public dataset, and 
Honolulu falls just below this threshold. However, it is the 
only MSA in Hawaii with a population greater than 500,000, so you can filter for it based on state and MSA size category.

```{r, message=FALSE}
hnlulu_trips <- here("nhts",
                     "trippub.csv") %>%
  read_csv() %>%
  filter(HHSTATE == "HI" &
           MSASIZE == "03")
```

# Filter out non-motorized trips

If your model does not include walk and bike trips, you should
filter these out.

```{r}
hnlulu_trips <- hnlulu_trips %>%
  filter(TRPTRANS != "01" & # Walk
           TRPTRANS != "02" & # Bike
           TRPTRANS != "19") # Airplane
```


# Count trips by trip purpose

Your model should include the following trip purposes:

* Home-based work (HBW)
* Home-based other (HBO)
* Non-home-based (NHB)

Classify trips into one of these three categories.

```{r}
hnlulu_trips <- hnlulu_trips %>%
  mutate(home_based = (WHYFROM == "01" |
                         WHYFROM == "02" |
                         WHYTO == "01" |
                         WHYTO == "02"),
         work = (WHYFROM == "03" |
           WHYTO == "03")) %>%
  mutate(purpose = case_when(home_based & work ~ "HBW",
                             home_based ~ "HBO",
                             TRUE ~ "NHB"))
```

You can quickly check how many of the total trips in your
study area fall into each category.

```{r}
table(hnlulu_trips$purpose)
```

In this example, we see that just over half of all trips are
in the "home-based other" category. Now we want to see how
many trips of each type each household makes.

```{r}
hh_trip_counts <- hnlulu_trips %>%
  group_by(HOUSEID) %>%
  summarise(HBO_trips = sum(purpose == "HBO"),
            HBW_trips = sum(purpose == "HBW"),
            NHB_trips = sum(purpose == "NHB"))
```

# Load household data

The trips file includes most of the household-level data you'd
need, but it won't include households that made no survey-day
trips, so it's better to get this by joining your trip counts
to the household file.

First you can load the household data and filter it the same way
you did for the trips. Then you can join the trip counts. 
Households with no trips won't have been in the trips file, so
you can replace the resulting NA values with zeros.

```{r, message=FALSE}
hnlulu_hhs <- here("nhts",
                     "hhpub.csv") %>%
  read_csv() %>%
  filter(HHSTATE == "HI" &
           MSASIZE == "03") %>%
  left_join(hh_trip_counts) %>%
  replace_na(list(HBO_trips = 0,
                  HBW_trips = 0,
                  NHB_trips = 0))
```

# Select (and create) variables

Spend some time browsing through the codebook and you'll find 
a lot of interesting variables that might be interesting 
predictors of trip generation, or travel behvior more generally.

You'll need to limit yourself to variables that have zone-level
analogs you can get from the census.

Let's say the data you've collected about each zone includes:

* Median income
* Number of zero-vehicle households
* Number of one-vehicle households
* Number of households with two or more vehicles
* Number of households with children

You would want to estimate a model that predicts the number of
household trip productions for each trip purpose based on 

* Household income
* The presence of a household vehicle
* The presence of household children

The NHTS presents income in one of eleven categories. I'll use 
the mid-point of each category to generate a continuous variable
indicating the household income in thousands. I'll have to make 
a guess for the midpoint of the top category ($200,000 or more).
I'm just going to use $300,000 - but this is a totally arbitrary
decision.

```{r}
hnlulu_hhs <- hnlulu_hhs %>%
  mutate(inc_k = case_when(HHFAMINC == "01" ~ 5,
                           HHFAMINC == "02" ~ 12.5,
                           HHFAMINC == "03" ~ 17.5,
                           HHFAMINC == "04" ~ 20,
                           HHFAMINC == "05" ~ 37.5,
                           HHFAMINC == "06" ~ 62.5,
                           HHFAMINC == "07" ~ 82.5,
                           HHFAMINC == "08" ~ 112.5,
                           HHFAMINC == "09" ~ 132.5,
                           HHFAMINC == "10" ~ 175,
                           HHFAMINC == "11" ~ 300,
                           TRUE ~ 99)) %>%
  mutate(inc_k = na_if(inc_k, 99))
```

And I'll create a categorical variable for vehicle ownership.

```{r}
hnlulu_hhs <- hnlulu_hhs %>%
  mutate(veh_cat = case_when(HHVEHCNT == 0 ~ "zero",
                             HHVEHCNT == 1 ~ "one",
                             TRUE ~ "two-plus"))
```

And a binary variable for whether there are children in the
household.

```{r}
hnlulu_hhs <- hnlulu_hhs %>%
  mutate(any_kids = LIF_CYC != "01" &
                    LIF_CYC != "02" &
                    LIF_CYC != "09" &
                    LIF_CYC != "10")
```

# Estimate a linear regression model

Now you can estimate a linear regression model for each trip 
purpose. I'm log-transforming income because I expect that
a percent. I'm using a base-two log so that I can interpret 
the coefficient as the effect of doubling income.

```{r}
hbo_trip_prod <- lm(HBO_trips ~ log2(inc_k) +
                                veh_cat +
                                any_kids,
                    data = hnlulu_hhs)
```

And here are the model results:

```{r}
summary(hbo_trip_prod)
```

The intercept is 1.18. This is the predicted number
of daily trips for a household with an annual income of $1,000
per year (so the base-two logarithm of the income, in thousands,
is zero), with one vehicle and no children.

A doubling of income is associated with 0.17 additional 
"home-based other" trips per day, although this effect is not
significant at a 95-percent (or even a 65-percent) confidence
level. 

Households with two or more vehicles make 1.3 more trips per day
than households with one vehicle. Households with no vehicles 
make fewer trips per day than those with one vehicle, but this 
difference is not significant at even an 80 percent confidence
level.

Households with children make, on average, 2.5 more daily trips
than those without children, holding income and vehicle access
constant.

What would be the predicted number of daily home-based other 
trips for a household with an annual income of $32,000 per 
year, two vehicles, and children living at home?

32 is $2^5$, so, this would be:

$1.18 + (5\times0.17) + 1.34 + 2.5$ or 5.9 trips.

This is the number of _trips produced_ by that household, or 
number of daily trip productions.
