---
title: "Skills for Analysis Assignment 9"
author: "GSD SES 5394"
date: "Spring 2023"
output: 
  rmdformats::downcute
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

The following methods will be useful for Analysis Assignment 9.

# Required packages

The skills in this tutorial draw on the `here` package, the `tidyverse` package, the `tigris` package, and the `sf` package, which are all familiar to you now.

```{r, warning=FALSE, message=FALSE, results='hide'}
library(tidyverse)
library(sf)
library(here)
library(tigris)
```

# Calculate total vehicle trips for each PA pair

You should have finished Assignment 8 with a table of production-attraction
flows, including the following columns:

* Production zone GEOID
* Attraction zone GEOID
* Number of HBO trips by SOV
* Number of HBO trips by HOV
* Number of HBW trips by SOV
* Number of HBW trips by HOV
* Number of NHB trips by SOV
* Number of NHB trips by HOV

You can use combine these columns to calculate the total number of
SOV trips for each trip purpose, and the total number of HOV trips for 
each trip purpose. You can get the number of HOV _vehicle_ trips by 
dividing the number of HOV (person) trips by an assumed vehicle occupancy
(this should be consistent with the vehicle occupancy you used for calculating
costs in the previous assignment). The number of SOV vehicle trips is the
same as the number of SOV person trips.

Finally add up all vehicle trips across both modes and all three trip 
purposes to get the total number of vehicles traveling between each 
production-attraction pair.

# Prepare PA matrix for import to TransCAD

You'll end up with a table that looks something like this:

```{r, echo=FALSE, message=FALSE, results='hide'}
PA_mat = here("Examples",
              "OKC",
              "pa-mat-ctv.csv") %>%
  read_csv(show_col_types = FALSE)

id_geoid_xwalk <- here("Examples",
                      "OKC",
                      "zones.geojson") %>%
  st_read() %>%
  st_drop_geometry() %>%
  select(GEOID, centroid)

xwalk_from <- id_geoid_xwalk %>%
  rename(from_node = centroid,
         from_GEOID = GEOID)

xwalk_to <- id_geoid_xwalk %>%
  rename(to_node = centroid,
         to_GEOID = GEOID)

PA_mat_GEOID <- left_join(PA_mat, xwalk_from) %>%
  left_join(xwalk_to) %>%
  select(from_GEOID, to_GEOID, flow) %>%
  replace_na(list(flow = 0))

```

```{r}
head(PA_mat_GEOID)
```

You'll want to change the GEOID values to the node IDs that
TransCAD is using.

## Export centroids with node IDs from TransCAD

Here is a video that demonstrates how to export the centroid
locations with ID numbers from TransCAD:

[Exporting TransCAD centroids with IDs](https://harvard.zoom.us/rec/share/2zTI_LGWk5ltj0ZcwhVctngImhnPMfHl9G7fhkto4lqdFOP_TBIZHJRLnf6RwSR-.Bhi2l7A9LRboEkQr?startTime=1680106518000){target="_blank"}

# Add centroid node IDs to PA matrix

Now you can take those centroid locations you just exported from TransCAD
and open them in R.

```{r, results='hide', message=FALSE}
transcad_points <- here("Examples",
                  "OKC",
                  "okc-centroids.json") %>%
  st_read()
```

Let's take a look at the first few rows:

```{r}
head(transcad_points)
```

## Associate PA matrix with TransCAD IDs

Now we want to associate each of those points from TransCAD with
a GEOID. I'll get the locations of the census tracts from the
`tigris` package. I want to just use the tract centroids, and 
I need everything to be in the WGS 84 coordinate reference system.

```{r, message=FALSE, results='hide'}
GEOID_tracts <- tracts(state = "OK",
                       county = c("Canadian",
                          "cleveland",
                          "Grady",
                          "Lincoln",
                          "Logan",
                          "McClain",
                          "Oklahoma")) %>%
  st_transform("WGS84") %>%
  st_centroid()
```

Now I can use the st_join to associate TransCAD's ID numbers
with the GEOID for each tract.

```{r}
xwalk <- st_join(GEOID_tracts, transcad_points, 
                 join = st_nearest_feature) %>%
  select(GEOID, ID_copy) %>%
  st_drop_geometry()

head(xwalk)
```

And from there, I can switch out the GEOIDs in my PA matrix
for the corresponding TransCAD IDs.

```{r}
xwalk_from <- xwalk %>%
  rename(from_node = ID_copy,
         from_GEOID = GEOID)

xwalk_to <- xwalk %>%
  rename(to_node = ID_copy,
         to_GEOID = GEOID)

PA_mat_transcad <- left_join(PA_mat_GEOID, xwalk_from) %>%
  left_join(xwalk_to) %>%
  select(from_node, to_node, flow)

head(PA_mat_transcad)
```

Now you can save that as a csv file that you'll open in TransCAD.

```{r}
write_csv(PA_mat_transcad,
          here("Examples",
               "OKC",
               "PA_mat_total_vehs.csv"))
```


# Import PA matrix to TransCAD

Here's a video that demonstrates how to open that as a
production-attraction matrix in TransCAD.

[Import PA matrix into TransCAD](https://harvard.zoom.us/rec/share/vQYyxjyqmBJuNBqkEefK6vncQqZW6KXyQgnB9dRfSG9QoEVWuZJCut-DQhOWTlR7.9KtpLmldQhGCHkcI?startTime=1680108053000){target="_blank"}

# Generate 1-hour OD ma

You'll need to generate a one-hour origin-destination matrix from a
24-hour production-attraction matrix.

[Generate OD matrix from PA matrix](https://harvard.zoom.us/rec/share/3ckEyAjUBm5CGOVDZxtZuVWvLcdsV5xGuwfZU82YHYDXd0ti_tO6Y4Q-ow6bWzUi.YUn5sGBOB8etfAYG?startTime=1680108237000){target="_blank"}

# Add capacities to road network

The trip assignment method requires you to define the capacity
of each roadway link and recreate the newtwork file.

[Define capacities](https://harvard.zoom.us/rec/share/jnwAY-tR7t-TP1Cpl_I4SDIVh2Ahzt0YUSb0RmwQ6kX6sBzDZuabogjDH4Fl8CaM.0oxWTsZgsgv7JS9_?startTime=1680108484000){target="_blank"}

# Estimate trip assignment

And now you can assign trips to the network.

[Assign trips](https://harvard.zoom.us/rec/share/4NlMcM0HBDow1duAtqD7LTQnF92haC5JgT56Qtvxuz_-DV_cAMdjEHoJw5yGBplk.FUZ7dCLPFyG02RD4?startTime=1680108824000){target="_blank"}

# Visualize link flows

Finally, you can visualize link flows within TransCAD.

[Visualize link flows](https://harvard.zoom.us/rec/share/KrUTFCNDAr4n86WEZI6uSMMRzh8zI7Ddbap4_nmQrvyGhnrDmjFvKwotDQY2wVlv.vbXTK6BHkgtop9YG?startTime=1680110777000){target="_blank"}

You can also export your layers and use R or ArcGIS to do further visualization.